A generalization of the Bradley-Terry model for draws in chess

### 2019-05-26 inst/collusion_chess_draws.Rnw
include pointer to github
various corrections suggested by Alna.  Thanks Alna\!

%% just as usual
\author{Robin K. S. Hankin\\Auckland University of Technology}
\title{A generalization of the Bradley-Terry model for draws in chess}

@@

\Keywords{Dirichlet distribution, hyperdirichlet distribution,
  combinatorics, \proglang{R}, multinomial distribution, constrained
optimization, Bradley-Terry}
\Plainkeywords{Dirichlet distribution, hyperdirichlet distribution,
  combinatorics, R, multinomial distribution, constrained
optimization, Bradley-Terry}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way competitions such as {\em MasterChef
  Australia}~\citep{hankin2017}, and non-competitive situations such
as consumer behaviour~\citep{hankin2010} and mating
preference~\citep{west2008}.

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage
winning about 37.5\% of games (black wins 27.6\%); and secondly, many
chess games, about 39.4\%, are drawn (figures from Chessbase).

\paragraph{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``white
to play and win''---to~\citet{adorjan1988}---``black is OK!''.

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citet{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

\section{Generalized Bradley-Terry}

Consider first the standard Bradley-Terry model for a finite number of
chess players.  Ignoring the effect of playing white, we wish to find
numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$ with
$p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating
$j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function,
on the assumption of independence of trials, would be

@@

\begin{equation}
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

@@

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis.  Unlike the present work,
which uses a dataset restricted to game result (win/loss/draw),
\citet{moul2009} included number of moves in each game\footnote{The
  reasoning was that hard-fought games that happened to end in a draw
  are more likely to last longer than collusive or pre-arranged
  draws.} in their analysis, and used a dataset of 7355 games played
in the period 1940-1978 to calibrate the ability of the players.  In
contrast, the analysis presented uses a dataset restricted to game
outcome (win-loss-draw) in the competition itself.

@@

Here I consider only the game results, reproduced in
table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

@@

>>>>

%% just as usual
\author{Robin K. S. Hankin\\Department of Mathematical Sciences\\Auckland University of Technology}
\title{A generalization of the Bradley-Terry model for draws in chess}

@@

\Keywords{Likelihood functions, method of support, \proglang{R},
  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
  Draws}
\Plainkeywords{Likelihood functions, method of support, R,
  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
  Draws}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative number $p_i$ to player
$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
modified and extended many times, for example to account for multi-way
competitions such as {\em MasterChef Australia}~\citep{hankin2017},
and non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).

\subsection{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citet{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

\subsection{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``white
to play and win''---to~\citet{adorjan1988}---``black is OK!''.

\subsection{Generalized Bradley-Terry}

Here I show how Bradley-Terry may be generalized ot account for the
existence of draws and the first-mover advantage.  Consider first the
standard Bradley-Terry model for a finite number of chess players.
Ignoring the effect of playing white, we wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

@@

\begin{equation}
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

@@

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis based on number of moves
in each game\footnote{The reasoning was that hard-fought games that
  happened to end in a draw are more likely to last longer than
  collusive or pre-arranged draws.}.  Their analysis incorporated a
dataset of 7355 games played in the period 1940-1978, used to
calibrate the ability of the players.  In contrast, the analysis
presented here uses only outcome of games (win-loss-draw) in the
competition itself.

@@

Here I consider game results from Cura\c{c}ao 1962, reproduced in
Table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

@@

\subsection*{Supplementary material}

R code for all statistical analysis given here is available at
\url{https://github.com/RobinHankin/hyper2}.
